Polynomial Time and Dependent Types

### 2023-11-14 paper/paper-extended.tex
[ paper ] final edits

@@

Following these ideas, let us assume that the input is a natural
number, so we assume that there is some type of natural numbers
$\tyNat$. For point (i), we must be able to iterate over these natural
numbers, so we use a linear iterator defined by this typing rule:
\begin{displaymath}
  \inferrule*
  {\vdash M_z : A \\ x : A \vdash M_s : A \\ \Gamma \vdash N : \tyNat}
  {\Gamma \vdash \tmRec\,N\,\{\conZero \mapsto M_z; \conSucc(x) \mapsto M_s\} : A}
\end{displaymath}
Note that in the zero, $M_z$, and sucessor, $M_s$, cases, the context
is empty to ensure that these cases may be invoked as many times as
required. Point (ii) above is automatically satisfied by being in a
typed $\lambda$-calculus, where it is difficult to stop functions from
being composable. For point (iii), the iterator as given does not
allow us to nest iterations. Once the natural number input $n$ has
been used for an iteration, the linear typing discipine prevents us
from using it again (note the two separate contexts $\Gamma_1$,
$\Gamma_2$ in the rule for application). In order to allow nested
iterations, we add an operator to duplicate numbers:
\begin{displaymath}
  \inferrule*
  {\Gamma \vdash M : \tyNat}
  {\Gamma \vdash \dupNat\,M : \tyNat \otimes \tyNat}
\end{displaymath}
Somewhat surprisingly, this system is now sound and complete for
polynomial time. Crucially, this depends on the two things we have
\emph{not} allowed. First, we have disallowed the construction of new
natural numbers via the $\conZero$ and $\conSucc$
constructors\footnote{Actually, $\conZero$ would be acceptable, as
  well as any constant natural number. It is only unrestricted use of
  $\conSucc$ that is dangerous.}. If we were to permit this, then we
could use iteration over the input to construct addition,
multiplication (by repeated addition), and then exponentials (by
repeated multiplication). We therefore refer to this system as the
\emph{Cons-Free} system.  Because we cannot construct values of type
$\tyNat$ within the system, complete programs in this system are open
terms as we explained at the start of this section.

@@

We will see in \autoref{sec:consfree-sound} that this system is sound
for polytime by a realisability argument. Completeness can be seen
more directly by constructing a function that iterates a function for
a statically known polynomial number of times in the size of the
input. Assume that we have a known polynomial
$p(n) = c_dn^d + \dots + c_0$ of degree $d$ with natural number
coefficients and some single step function
$f : \mathit{St} \multimap \mathit{St}$ over a state type
$\mathit{St}$ that runs to completion for input of size $n$ in $p(n)$
steps. Then, using the iterator above we can iterate $f$ over a
$\tyNat$ representing the size of the input:
\begin{displaymath}
  \begin{array}{l}
    I_1 : \tyNat \multimap \mathit{St} \multimap \mathit{St} \\
    I_1 = \lambda n. \lambda s.\tmRec\,n\,\{\conZero \mapsto s; \conSucc(s) \mapsto f\,s \}
  \end{array}
\end{displaymath}
To achieve higher degrees, we can use $\dupNat$ to nest iterations:
\begin{displaymath}
  \begin{array}{l}
    I_{k+1} : \tyNat \multimap \mathit{St} \multimap \mathit{St} \\
    I_{k+1} = \lambda n. \lambda s.
    \begin{array}[t]{@{}l}
      \Let\,(n,n') = \dupNat\,n\,\In\\
      \tmRec\,n\,\{\conZero \mapsto s; \conSucc(s) \mapsto I_k\,n'\,s \}
    \end{array}
  \end{array}
\end{displaymath}
By further use of $\dupNat$ and composition to handle addition of
polynomials, the function $f$ can now be iterated $p(n)$ many times,
where $n$ is the input $\tyNat$. Thus, the \emph{Cons-free} system can
represent all polytime computations.

@@

The Cons-free system is sound and complete for polytime, but is quite
awkward from the point of view of functional programming. It allows us
to iterate over natural numbers that come from the input but does not
allow us to build further values to do iteration on. For example, if
our input is a list, then we cannot transform it into a binary search
tree and then flatten it, we must always refer back to the original
natural number input. Even dividing the input into two halves to be
treated separately is difficult.

@@

  \inferrule*
  {\Gamma_1 \vdash M : \Diamond \\ \Gamma_2 \vdash N : \tyNat}
  {\Gamma_1, \Gamma_2 \vdash \conSucc(M, N) : \tyNat}
\end{mathpar}
To construct a $\conZero$, we must have a $\Diamond$ to pay for it,
and likewise, to construct a $\conSucc$ we must pay a $\Diamond$. We
can think of $\Diamond$s as an unit of iterable data. Iterability is
``saved up'' in data during construction, and released during
iteration. Diamonds cannot be created by a program itself, for the
same reason that constructors were prohibited in the \emph{Cons-free}
system, but they are released from iterable data during iteration. The
LFPL natural number iterator has the following typing rule:
\begin{displaymath}
  \inferrule*
  {d : \Diamond \vdash M_z : A \\ d : \Diamond, x : A \vdash M_s : A \\ \Gamma \vdash N : \tyNat}
  {\Gamma \vdash \tmRec\,N\,\{\conZero(d) \mapsto M_z; \conSucc(d,x) \mapsto M_s \} : A}
\end{displaymath}
The difference with the \emph{Cons-free} iterator above is that the
$\conZero$ and $\conSucc$ cases now both have an additional binding of
type $\Diamond$. This allows some form nesting of iterations: during
an iteration over the input, the program can accumulate $\Diamond$s to
use for iteration over substructures that are smaller than the current
point in the iteration. A construction, due to \citet{AehligS02},
illustrates how this leads to all polytime computations. As above, we
assume a polynomial $p(n)$ and a step function
$f : \mathit{St} \multimap \mathit{St}$ that needs to be iterated
$p(n)$ times. We construct a linear iterator:
\begin{displaymath}
  \begin{array}{l}
    I_1 : (\tyNat \otimes \mathit{St}) \multimap (\tyNat \otimes \mathit{St}) \\
    I_1 = \lambda (n, s).\,\tmRec\,n\,\{
    \begin{array}[t]{@{}lcl}
      \conZero(d)&\mapsto&(\conZero(d), s);\\
      \conSucc(d,(n,s)) &\mapsto& (\conSucc(d,n),f\,s) \}
    \end{array}
  \end{array}
\end{displaymath}
Note that this iterator returns the natural number input as well as
the new state. LFPL does not allow duplication of iterable inputs, so
we must always reconstruct it if we want to do further
iteration. Addition of polynomials is accomplished by composition of
iterators. To raise the degree, we again use a nesting iterator:
\begin{displaymath}
  \begin{array}{l}
    I_{k+1} : (\tyNat \otimes \mathit{St}) \multimap (\tyNat \otimes \mathit{St}) \\
    I_{k+1} = \lambda (n,s).\,\tmRec\,n\,\{
    \begin{array}[t]{@{}lcl}
      \conZero(d)&\mapsto&(\conZero(d),s); \\
      \conSucc(d,(n,s)) &\mapsto& \Let\,(n,s)=I_k\,(n,s)\,\In\,(\conSucc(d,n), s) \}
    \end{array}
  \end{array}
\end{displaymath}
Unlike in the \emph{Cons-free} system, this iterator does not raise
the degree of the nested iterator directly. Rather, the iterator $I_k$
on the input $n$ performs $\binom{n}{k}$ iterations. As observed by
Aehlig and Schwictenberg, this is sufficient because the binomials
form a basis for the vector space of all polynomials.

@@

We have now seen the \emph{Cons-free} and \emph{LFPL} systems for
capturing polytime by means of linear typing and restricted
iteration. We now look to extend these systems to include dependent
types by building upon \emph{Quantitative Type Theory} (QTT)
\cite{atkey18qtt,mcbride16}. This section reviews QTT and describes
how we have adapted it to the polytime systems we saw in the previous
section.

@@

\subsubsection{Data Types} QTT, as we have presented it so far, has no
interesting base types to perform computation on. Following our
presentation of the simply typed linear systems in
\autoref{sec:linear-intro}, we add two kinds of datatype to
QTT. First, in \autoref{sec:noniter-qtt}, we add non-iterable
datatypes that allow construction and case analysis, but no
recursion. In \autoref{sec:cons-free-qtt} we describe how to extend
QTT to be a dependently typed adaptation of the Cons-free system of
\autoref{sec:cons-free-intro}. In \autoref{sec:lfpl-qtt}, we apply the
same treatment to the LFPL-style system.

@@

  \inferrule*
  {\Gamma_1 \vdash M \stackrel\sigma: T \\
    \Gamma_2 \vdash N \stackrel\sigma: \ListTy(T) \\
    0\Gamma_1 = 0\Gamma_2}
  {\Gamma_1 + \Gamma_2 \vdash \cCons(M,N) \stackrel\sigma: \ListTy(T)}
\end{mathpar}
Lists do have the potential for iteration by their recursive nature,
but in order to ensure the complexity guarantees we only permit
matching without recursion in the full calculus. Here is the rule for
dependently typed case analysis on lists, which also obeys the usual
$\beta$-equalities for case analysis, analogous to the ones for
Booleans:
\begin{mathpar}
  \inferrule*
  {0\Gamma_1, x \stackrel0: \ListTy(T) \vdash P~\istype \\
    \Gamma_1 \vdash M \stackrel\sigma: \ListTy(T) \\
    \Gamma_2 \vdash N_1 \stackrel\sigma: P[\cNil/x] \\
    \Gamma_2, h \stackrel\sigma: T, t \stackrel\sigma: \ListTy(T) \vdash N_2 \stackrel\sigma: P[\cCons(h,t)/x] \\
    0\Gamma_1 = 0\Gamma_2}
  {\Gamma_1 + \Gamma_2 \vdash \Match_{x.P}\,M\,\{\,\cNil \mapsto N_1; \cCons(h,t) \mapsto N_2\,\} \stackrel\sigma: P[M/x]}
\end{mathpar}
In the $\sigma = 0$ fragment, however, we are free to iterate on lists
because computations in this fragment are only meant for type-level
computation, not for the program itself. Put another way, the type
checker may perform arbitary recursion on lists to type check the
program, but the program itself may not do so without correctly
accounting its costs as described in the following sections. The
$\sigma = 0$ fragment recursor for lists has the following typing
rule, which is the standard dependent eliminator for lists, with
everything annotated as $0$ usage.
\begin{mathpar}
  \inferrule*
  {0\Gamma, x \stackrel0: \ListTy(T) \vdash P~\istype \\
    0\Gamma \vdash M \stackrel0: \ListTy(T) \\
    0\Gamma \vdash N_1 \stackrel0: P[\cNil/x] \\
    0\Gamma, h \stackrel0: T, t \stackrel0: \ListTy(T), p \stackrel0: P[t/x] \vdash N_2 \stackrel0: P[\cCons(h,t)/x]}
  {0\Gamma \vdash \mathrm{recList}_{x.P}\,M\,\{\,\cNil \mapsto N_1; \cCons(h,t;p) \mapsto N_2\,\} \stackrel0: P[M/x]}
\end{mathpar}
This eliminator also obeys the usual $\beta$-equality laws for a list
eliminator, using the resource freedom of the $\sigma = 0$ fragment to
duplicate the $N_2$ term in the $\cCons$ case.

@@

The datatypes of the previous section still only allow us to write
programs in the $\sigma = 1$ fragment that are constant time in the
size of their input. As with the simply typed linear system, if we are
handed a list of an unknown length, we can only explore it to a fixed
depth, determined statically by the program. To write programs that do
work proportional to the size of their input, we need some form of
iterable datatype. In both our Cons-free and LFPL-style QTT systems,
we use a natural number datatype.

The cons-free system cannot allow the programmer to construct natural
numbers in the $\sigma = 1$ fragment, as this would violate the
complexity guarantees. However, we can use the flexibility of QTT to
allow free construction of naturals in the $\sigma = 0$ fragment,
which allows us to use natural numbers freely in types. Therefore, we
have the following introduction rules, only usable in the $\sigma = 0$
fragment:
\begin{mathpar}
  \inferrule*
  {\Gamma~\isctxt}
  {\Gamma \vdash \conZero \stackrel0: \tyNat}

  \inferrule*
  {\Gamma \vdash M \stackrel0: \tyNat}
  {\Gamma \vdash \conSucc(M) \stackrel0: \tyNat}
\end{mathpar}
The cons-free system allows free duplication of complete natural
numbers. This is accomplished by a special construct copying the
simply linear typed rule we gave above:
\begin{mathpar}
  \inferrule*
  {\Gamma \vdash M \stackrel\sigma: \tyNat}
  {\Gamma \vdash \mathrm{dupNat}(M) \stackrel\sigma: \tyNat \otimes \tyNat}
\end{mathpar}
Anyone who has reasoned about the metatheory of, or implemented a type
checker for, dependent types will view this rule with unease as it
appears to grant the ability to construct non-canonical values of pair
type, and consequently generate non-canonical naturals. We fix this by
adding an equational rule to the $\sigma = 0$ fragment, ensuring
\emph{definitionally} that $\mathrm{dupNat}$ acts as its name implies:
\begin{mathpar}
  \inferrule*
  {\Gamma \vdash M \stackrel0: \tyNat}
  {\Gamma \vdash \mathrm{dupNat}(M) \equiv (M,M) \stackrel0: \tyNat \otimes \tyNat}
\end{mathpar}
Note that this rule is well-typed by the $0$-needs-$0$ property of
QTT, and the fact that $0+0=0$.

@@

As explained in \autoref{sec:lfpl-intro}, the LFPL system differs from
the Cons-free system in that it is possible to construct natural
numbers (and other iterable datatypes), provided one has the necessary
diamonds to pay for the construction. As with the natural number type
in the Cons-free system, it ought not be possible to construct
diamonds in the $\sigma = 1$ fragment, as this would amount to the
free distribution of diamonds to all which would lead to a collapse in
the complexity guarantees of the system. It is possible construct
diamonds in the $\sigma = 0$, though:
\begin{mathpar}
  \inferrule*
  {\Gamma~\isctxt}
  {0\Gamma \vdash \Diamond~\istype}

@@

  \inferrule*
  {\Gamma_1 \vdash M \stackrel\sigma: \Diamond \\
    \Gamma_2 \vdash N \stackrel\sigma: \tyNat \\
    0\Gamma_1 = 0\Gamma_2}
  {\Gamma_1 + \Gamma_2 \vdash \conSucc(M,N) \stackrel\sigma: \tyNat}
\end{mathpar}
In the $\sigma = 0$ fragment, we can construct $\Diamond$s for free,
and so construct natural numbers freely as well just as we did for the
Cons-free system above.

@@

Unlike the natural number iterator in the Cons-free system, this
iterator has $\beta$-equalities in both the $\sigma = 0$ and
$\sigma = 1$ fragments. In the $\conSucc$ case, for example, we have:
\begin{displaymath}
  \begin{array}{cl}
    & \tmRec\,(\conSucc(M_d,M_n))\,\{\conZero(d) \mapsto N_z; \conSucc(d,n;p) \mapsto N_s\} \\
    \equiv & N_s[M_d/d, M_n/n, \tmRec\,M_n\,\{\conZero(d) \mapsto N_z; \conSucc(d,n;p) \mapsto N_s\}/p]
  \end{array}
\end{displaymath}
Note that the fact that the variable $n$ in the $N_s$ term is
annotated $0$, which allows us to duplicate it even when we are in the
$\sigma = 1$ fragment.

Just as for the Cons-free system iterator above, in the $\sigma = 0$
fragment this rule is identical to the usual dependently typed
recursor for the natural numbers, so it can be used in the types to
reason about programs. Moreover, we will see in
\autoref{sec:lfpl-sound} that this rule is also sound for polynomial
time in a system with $\Diamond$s.

@@

We have only defined an iterable natural number datatype for both of
our systems above. We could extend both systems to include further
iterable inductive types, although in the \emph{Cons-free} system this
is not particularly useful due to the prohibition of
construction. However, sticking with just the natural numbers, we can
use the power of dependent types with a universe to create further
datatypes whose size is measured by some iterable natural
number. Iteration on the size yields iteration over the full
datastructure. For example, in the LFPL system, we can define a type
of iterable lists by pairing a size with a type of elements defined by
recursion on the size:
\begin{displaymath}
  \mathrm{IList}\,A = (n \stackrel1: \tyNat) \otimes \left(\mathrm{rec}_{x.\mathsf{U}}\,n\,\{ \conZero(d) \mapsto I; \conSucc(d,n;p) \mapsto A \otimes p \}\right)
\end{displaymath}
The $\cNil$ and $\cCons$ constructors can now be defined in terms of
$\conZero$ and $\conSucc$, provided the caller supplies sufficient
$\Diamond$s. These definitions live in the $\sigma = 1$ fragment, so
we annotate them appropriately:
\begin{mathpar}
  \begin{array}{l}
    \mathrm{nil} \stackrel1: \Diamond \to \mathrm{IList}\,A \\
    \mathrm{nil}\,d = (\conZero(d), *)
  \end{array}

@@

We can use the reflection type former defined in
\autoref{sec:qtt-reflection} to define a predicate on decision
problems that establishes whether or not they are polytime decision
problems. Specifically, we can state that there is a polytime
realisable boolean-value predicate that reports true exactly when the
given element of $a$ is in the predicate:
\begin{displaymath}
  \mathrm{PTIME}(A,P) =
    (f \stackrel1: \mathbf{R}(A \to \BoolTy)) \otimes (a \stackrel1: A) \to (\mathbf{R}^{-1}(f)\, a = \cTrue) \Leftrightarrow P\,a
\end{displaymath}
Thus, $\mathrm{PTIME}(A,P)$ is a logical proposition stating that the
decision problem $(A,P)$ is decidable in polytime. We make three notes
about this definition: (i) \emph{proofs} of $\mathrm{PTIME}(A,P)$, are
carried out in the $\sigma = 0$ fragment, where we have the full power
of Type Theory to aid us; (ii) this definition is intrinsic, in the
sense that, whichever of the polytime systems is chosen, proving that
a decision problem is solvable in polytime is a matter of programming,
without having to reason directly about machine models and step
counting; and (iii) we have defined problems to have arbitrary types
$A$ as domains, rather than bitstrings, and so the notion of size
attached to an input is intrinsic to the type $A$ chosen.

>>>>

\newcommand{\ConsFree}{\emph{Cons-free}}

@@

Following these ideas, let us assume that the input is a natural
number, so we assume that there is some type of natural numbers
$\tyNat$. For point (i), we must be able to iterate over these natural
numbers, so we use a linear iterator defined by this typing rule:
\begin{displaymath}
  \inferrule*
  {\vdash M_z : A \\ x : A \vdash M_s : A \\ \Gamma \vdash N : \tyNat}
  {\Gamma \vdash \tmRec\,N\,\{\conZero \mapsto M_z; \conSucc(x) \mapsto M_s\} : A}
\end{displaymath}
Note that in the zero, $M_z$, and sucessor, $M_s$, cases, the context
is empty to ensure that these cases may be invoked as many times as
required. Point (ii) above is automatically satisfied by being in a
typed $\lambda$-calculus, where it is difficult to stop functions from
being composable. For point (iii), the iterator as given does not
allow us to nest iterations. Once the natural number input $n$ has
been used for an iteration, the linear typing discipine prevents us
from using it again (note the two separate contexts $\Gamma_1$,
$\Gamma_2$ in the rule for application). In order to allow nested
iterations, we add an operator to duplicate numbers:
\begin{displaymath}
  \inferrule*
  {\Gamma \vdash M : \tyNat}
  {\Gamma \vdash \dupNat\,M : \tyNat \otimes \tyNat}
\end{displaymath}
Somewhat surprisingly, this system is now sound and complete for
polynomial time. Crucially, this depends on the two things we have
\emph{not} allowed. First, we have disallowed the construction of new
natural numbers via the $\conZero$ and $\conSucc$
constructors\footnote{Actually, $\conZero$ would be acceptable, as
  well as any constant natural number. It is only unrestricted use of
  $\conSucc$ that is dangerous.}. If we were to permit this, then we
could use iteration over the input to construct addition,
multiplication (by repeated addition), and then exponentials (by
repeated multiplication). We therefore refer to this system as the
\ConsFree{} system.  Because we cannot construct values of type
$\tyNat$ within the system, complete programs in this system are open
terms as we explained at the start of this section.

@@

We will see in \autoref{sec:consfree-sound} that this system is sound
for polytime by a realisability argument. Completeness can be seen
more directly by constructing a function that iterates a function for
a statically known polynomial number of times in the size of the
input. Assume that we have a known polynomial
$p(n) = c_dn^d + \dots + c_0$ of degree $d$ with natural number
coefficients and some single step function
$f : \mathit{St} \multimap \mathit{St}$ over a state type
$\mathit{St}$ that runs to completion for input of size $n$ in $p(n)$
steps. Then, using the iterator above we can iterate $f$ over a
$\tyNat$ representing the size of the input:
\begin{displaymath}
  \begin{array}{l}
    I_1 : \tyNat \multimap \mathit{St} \multimap \mathit{St} \\
    I_1 = \lambda n. \lambda s.\tmRec\,n\,\{\conZero \mapsto s; \conSucc(s) \mapsto f\,s \}
  \end{array}
\end{displaymath}
To achieve higher degrees, we can use $\dupNat$ to nest iterations:
\begin{displaymath}
  \begin{array}{l}
    I_{k+1} : \tyNat \multimap \mathit{St} \multimap \mathit{St} \\
    I_{k+1} = \lambda n. \lambda s.
    \begin{array}[t]{@{}l}
      \Let\,(n,n') = \dupNat\,n\,\In\\
      \tmRec\,n\,\{\conZero \mapsto s; \conSucc(s) \mapsto I_k\,n'\,s \}
    \end{array}
  \end{array}
\end{displaymath}
By further use of $\dupNat$ and composition to handle addition of
polynomials, the function $f$ can now be iterated $p(n)$ many times,
where $n$ is the input $\tyNat$. Thus, the \ConsFree{} system can
represent all polytime computations.

@@

The \ConsFree{} system is sound and complete for polytime, but is quite
awkward from the point of view of functional programming. It allows us
to iterate over natural numbers that come from the input but does not
allow us to build further values to do iteration on. For example, if
our input is a list, then we cannot transform it into a binary search
tree and then flatten it, we must always refer back to the original
natural number input. Even dividing the input into two halves to be
treated separately is difficult.

@@

  \inferrule*
  {\Gamma_1 \vdash M : \Diamond \\ \Gamma_2 \vdash N : \tyNat}
  {\Gamma_1, \Gamma_2 \vdash \conSucc(M, N) : \tyNat}
\end{mathpar}
To construct a $\conZero$, we must have a $\Diamond$ to pay for it,
and likewise, to construct a $\conSucc$ we must pay a $\Diamond$. We
can think of $\Diamond$s as an unit of iterable data. Iterability is
``saved up'' in data during construction, and released during
iteration. Diamonds cannot be created by a program itself, for the
same reason that constructors were prohibited in the \ConsFree{}
system, but they are released from iterable data during iteration. The
LFPL natural number iterator has the following typing rule:
\begin{displaymath}
  \inferrule*
  {d : \Diamond \vdash M_z : A \\ d : \Diamond, x : A \vdash M_s : A \\ \Gamma \vdash N : \tyNat}
  {\Gamma \vdash \tmRec\,N\,\{\conZero(d) \mapsto M_z; \conSucc(d,x) \mapsto M_s \} : A}
\end{displaymath}
The difference with the \ConsFree{} iterator above is that the
$\conZero$ and $\conSucc$ cases now both have an additional binding of
type $\Diamond$. This allows some form nesting of iterations: during
an iteration over the input, the program can accumulate $\Diamond$s to
use for iteration over substructures that are smaller than the current
point in the iteration. A construction, due to \citet{AehligS02},
illustrates how this leads to all polytime computations. As above, we
assume a polynomial $p(n)$ and a step function
$f : \mathit{St} \multimap \mathit{St}$ that needs to be iterated
$p(n)$ times. We construct a linear iterator:
\begin{displaymath}
  \begin{array}{l}
    I_1 : (\tyNat \otimes \mathit{St}) \multimap (\tyNat \otimes \mathit{St}) \\
    I_1 = \lambda (n, s).\,\tmRec\,n\,\{
    \begin{array}[t]{@{}lcl}
      \conZero(d)&\mapsto&(\conZero(d), s);\\
      \conSucc(d,(n,s)) &\mapsto& (\conSucc(d,n),f\,s) \}
    \end{array}
  \end{array}
\end{displaymath}
Note that this iterator returns the natural number input as well as
the new state. LFPL does not allow duplication of iterable inputs, so
we must always reconstruct it if we want to do further
iteration. Addition of polynomials is accomplished by composition of
iterators. To raise the degree, we again use a nesting iterator:
\begin{displaymath}
  \begin{array}{l}
    I_{k+1} : (\tyNat \otimes \mathit{St}) \multimap (\tyNat \otimes \mathit{St}) \\
    I_{k+1} = \lambda (n,s).\,\tmRec\,n\,\{
    \begin{array}[t]{@{}lcl}
      \conZero(d)&\mapsto&(\conZero(d),s); \\
      \conSucc(d,(n,s)) &\mapsto& \Let\,(n,s)=I_k\,(n,s)\,\In\,(\conSucc(d,n), s) \}
    \end{array}
  \end{array}
\end{displaymath}
Unlike in the \ConsFree{} system, this iterator does not raise
the degree of the nested iterator directly. Rather, the iterator $I_k$
on the input $n$ performs $\binom{n}{k}$ iterations. As observed by
Aehlig and Schwictenberg, this is sufficient because the binomials
form a basis for the vector space of all polynomials.

@@

We have now seen the \ConsFree{} and LFPL systems for
capturing polytime by means of linear typing and restricted
iteration. We now look to extend these systems to include dependent
types by building upon \emph{Quantitative Type Theory} (QTT)
\cite{atkey18qtt,mcbride16}. This section reviews QTT and describes
how we have adapted it to the polytime systems we saw in the previous
section.

@@

\subsubsection{Data Types} QTT, as we have presented it so far, has no
interesting base types to perform computation on. Following our
presentation of the simply typed linear systems in
\autoref{sec:linear-intro}, we add two kinds of datatype to
QTT. First, in \autoref{sec:noniter-qtt}, we add non-iterable
datatypes that allow construction and case analysis, but no
recursion. Then, in \autoref{sec:cons-free-qtt} we describe how to
extend QTT to be a dependently typed adaptation of the \ConsFree{}
system of \autoref{sec:cons-free-intro} by adding a type of iterable
naturals. In \autoref{sec:lfpl-qtt} we apply the same treatment to
the LFPL-style system.

@@

  \inferrule*
  {\Gamma_1 \vdash M \stackrel\sigma: T \\
    \Gamma_2 \vdash N \stackrel\sigma: \ListTy(T) \\
    0\Gamma_1 = 0\Gamma_2}
  {\Gamma_1 + \Gamma_2 \vdash \cCons(M,N) \stackrel\sigma: \ListTy(T)}
\end{mathpar}
Lists do have the potential for iteration by their recursive nature,
but in order to ensure the polytime complexity guarantees we only permit
matching without recursion in the $\sigma = 1$ fragment. Here is the rule for
dependently typed case analysis on lists, which also obeys the usual
$\beta$-equalities for case analysis, analogous to the ones for
booleans:
\begin{mathpar}
  \inferrule*
  {0\Gamma_1, x \stackrel0: \ListTy(T) \vdash P~\istype \\
    \Gamma_1 \vdash M \stackrel\sigma: \ListTy(T) \\
    \Gamma_2 \vdash N_1 \stackrel\sigma: P[\cNil/x] \\
    \Gamma_2, h \stackrel\sigma: T, t \stackrel\sigma: \ListTy(T) \vdash N_2 \stackrel\sigma: P[\cCons(h,t)/x] \\
    0\Gamma_1 = 0\Gamma_2}
  {\Gamma_1 + \Gamma_2 \vdash \Match_{x.P}\,M\,\{\,\cNil \mapsto N_1; \cCons(h,t) \mapsto N_2\,\} \stackrel\sigma: P[M/x]}
\end{mathpar}
In the $\sigma = 0$ fragment, however, we are free to iterate on lists
because computations in this fragment are only meant for type-level
computation, not for the program itself. Put another way, the type
checker may perform arbitary recursion on lists to type check the
program, but the program itself may not do so without correctly
accounting its costs as described in the following sections. The
$\sigma = 0$ fragment recursor for lists has the following typing
rule, which is the standard dependent eliminator for lists except that
everything annotated as $0$ usage.
\begin{mathpar}
  \inferrule*
  {0\Gamma, x \stackrel0: \ListTy(T) \vdash P~\istype \\
    0\Gamma \vdash M \stackrel0: \ListTy(T) \\
    0\Gamma \vdash N_1 \stackrel0: P[\cNil/x] \\
    0\Gamma, h \stackrel0: T, t \stackrel0: \ListTy(T), p \stackrel0: P[t/x] \vdash N_2 \stackrel0: P[\cCons(h,t)/x]}
  {0\Gamma \vdash \mathrm{recList}_{x.P}\,M\,\{\,\cNil \mapsto N_1; \cCons(h,t;p) \mapsto N_2\,\} \stackrel0: P[M/x]}
\end{mathpar}
This eliminator also obeys the usual $\beta$-equality laws for a list
eliminator, using the resource freedom of the $\sigma = 0$ fragment to
duplicate the $N_2$ term in the $\cCons$ case.

@@

The datatypes of the previous section still only allow us to write
programs in the $\sigma = 1$ fragment that are constant time in the
size of their input. As with the simply typed linear system, if we are
handed a list of an unknown length, we can only explore it to a fixed
depth, determined statically by the program. To write programs that do
work proportional to the size of their input, we need some form of
iterable datatype. In both our \ConsFree{} and LFPL-style QTT systems,
we use a natural number datatype.

The \ConsFree{} system cannot allow the programmer to construct natural
numbers in the $\sigma = 1$ fragment, as this would violate the
complexity guarantees. However, we can use the flexibility of QTT to
allow free construction of naturals in the $\sigma = 0$ fragment,
which allows us to use natural numbers freely in types. Therefore, we
have the following introduction rules, only usable in the $\sigma = 0$
fragment:
\begin{mathpar}
  \inferrule*
  {\Gamma~\isctxt}
  {\Gamma \vdash \conZero \stackrel0: \tyNat}

  \inferrule*
  {\Gamma \vdash M \stackrel0: \tyNat}
  {\Gamma \vdash \conSucc(M) \stackrel0: \tyNat}
\end{mathpar}
The cons-free system allows free duplication of complete natural
numbers. This is accomplished by a special construct copying the
simply linear typed rule we gave above:
\begin{mathpar}
  \inferrule*
  {\Gamma \vdash M \stackrel\sigma: \tyNat}
  {\Gamma \vdash \mathrm{dupNat}(M) \stackrel\sigma: \tyNat \otimes \tyNat}
\end{mathpar}
Anyone who has reasoned about the metatheory of, or implemented a type
checker for, dependent types will view this rule with unease as it
appears to grant the ability to construct non-canonical values of pair
type, and consequently generate non-canonical naturals. We fix this by
adding an equational rule to the $\sigma = 0$ fragment, ensuring
definitionally that $\mathrm{dupNat}$ acts as its name implies:
\begin{mathpar}
  \inferrule*
  {\Gamma \vdash M \stackrel0: \tyNat}
  {\Gamma \vdash \mathrm{dupNat}(M) \equiv (M,M) \stackrel0: \tyNat \otimes \tyNat}
\end{mathpar}
Note that this rule is well-typed by the $0$-needs-$0$ property of
QTT, and the fact that $0+0=0$.

@@

As explained in \autoref{sec:lfpl-intro}, the LFPL system differs from
the \ConsFree{} system in that it is possible to construct natural
numbers (and other iterable datatypes), provided one has the necessary
diamonds to pay for the construction. As with the natural number type
in the \ConsFree{} system, it ought not be possible to construct
diamonds in the $\sigma = 1$ fragment, as this would amount to the
free distribution of diamonds to all which would lead to a collapse in
the complexity guarantees of the system. It is possible construct
diamonds in the $\sigma = 0$, though:
\begin{mathpar}
  \inferrule*
  {\Gamma~\isctxt}
  {0\Gamma \vdash \Diamond~\istype}

@@

  \inferrule*
  {\Gamma_1 \vdash M \stackrel\sigma: \Diamond \\
    \Gamma_2 \vdash N \stackrel\sigma: \tyNat \\
    0\Gamma_1 = 0\Gamma_2}
  {\Gamma_1 + \Gamma_2 \vdash \conSucc(M,N) \stackrel\sigma: \tyNat}
\end{mathpar}
In the $\sigma = 0$ fragment, we can construct $\Diamond$s for free,
and so construct natural numbers freely as well just as we did for the
\ConsFree{} system above.

@@

Unlike the natural number iterator in the \ConsFree{} system, this
iterator has $\beta$-equalities in both the $\sigma = 0$ and
$\sigma = 1$ fragments. In the $\conSucc$ case, for example, we have:
\begin{displaymath}
  \begin{array}{cl}
    & \tmRec\,(\conSucc(M_d,M_n))\,\{\conZero(d) \mapsto N_z; \conSucc(d,n;p) \mapsto N_s\} \\
    \equiv & N_s[M_d/d, M_n/n, \tmRec\,M_n\,\{\conZero(d) \mapsto N_z; \conSucc(d,n;p) \mapsto N_s\}/p]
  \end{array}
\end{displaymath}
Note that the fact that the variable $n$ in the $N_s$ term is
annotated $0$, which allows us to use $M_n$ twice even when we are in
the $\sigma = 1$ fragment.

Just as for the \ConsFree{} system iterator above, in the $\sigma = 0$
fragment this rule is identical to the usual dependently typed
recursor for the natural numbers, so it can be used in the types to
reason about programs. Moreover, we will see in
\autoref{sec:lfpl-sound} that this rule is also sound for polynomial
time in a system with $\Diamond$s.

@@

We have only defined an iterable natural number datatype for both of
our systems above. We could extend both systems to include further
iterable inductive types, although in the \ConsFree{} system this
is not particularly useful due to the prohibition of
construction. However, sticking with just the natural numbers, we can
use the power of dependent types with a universe to create further
datatypes whose size is measured by some iterable natural
number. Iteration on the size yields iteration over the full
datastructure. For example, in the LFPL system, we can define a type
of iterable lists by pairing a size with a type of elements defined by
recursion on the size:
\begin{displaymath}
  \mathrm{IList}\,A = (n \stackrel1: \tyNat) \otimes \left(\mathrm{rec}_{x.\mathsf{U}}\,n\,\{ \conZero(d) \mapsto I; \conSucc(d,n;p) \mapsto A \otimes p \}\right)
\end{displaymath}
The $\cNil$ and $\cCons$ constructors can now be defined in terms of
$\conZero$ and $\conSucc$, provided the caller supplies sufficient
$\Diamond$s. These definitions live in the $\sigma = 1$ fragment, so
we annotate them appropriately:
\begin{mathpar}
  \begin{array}{l}
    \mathrm{nil} \stackrel1: \Diamond \to \mathrm{IList}\,A \\
    \mathrm{nil}\,d = (\conZero(d), *)
  \end{array}

@@

We can use the reflection type former defined in
\autoref{sec:qtt-reflection} to define a predicate on decision
problems that establishes whether or not they are polytime decision
problems. Specifically, we can state that there is a polytime
realisable boolean-value predicate that reports true exactly when the
given element of $a$ is in the predicate:
\begin{displaymath}
  \mathrm{PTIME}(A,P) =
    (f \stackrel1: \mathbf{R}(A \to \BoolTy)) \otimes \left((a \stackrel1: A) \to (\mathbf{R}^{-1}(f)\, a = \cTrue) \Leftrightarrow P\,a\right)
\end{displaymath}
Thus, $\mathrm{PTIME}(A,P)$ is a logical proposition stating that the
decision problem $(A,P)$ is decidable in polytime. We make three notes
about this definition: (i) \emph{proofs} of $\mathrm{PTIME}(A,P)$, are
carried out in the $\sigma = 0$ fragment, where we have the full power
of Type Theory to aid us; (ii) this definition is intrinsic, in the
sense that, whichever of the polytime systems is chosen, proving that
a decision problem is solvable in polytime is a matter of programming,
without having to reason directly about machine models and step
counting; and (iii) we have defined problems to have arbitrary types
$A$ as domains, rather than bitstrings, and so the notion of size
attached to an input is intrinsic to the type $A$ chosen.
